---
title: "Homework 2"
author: "Brent Lewis"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
---
---

```{r}
#| label: load-libraries
#| echo: false # This option disables the printing of code (only output is displayed).
#| message: false
#| warning: false

library(tidyverse)
library(wbstats)
library(skimr)
library(countrycode)
library(here)
```

# Mass shootings in the US

```{r}
#| echo: false
#| message: false
#| warning: false


mass_shootings <- read_csv(here::here("data", "mass_shootings.csv"))

glimpse(mass_shootings)
```

| column(variable)     | description                                                                 |
|--------------------------|----------------------------------------------|
| case                 | short name of incident                                                      |
| year, month, day     | year, month, day in which the shooting occurred                             |
| location             | city and state where the shooting occcurred                                 |
| summary              | brief description of the incident                                           |
| fatalities           | Number of fatalities in the incident, excluding the shooter                 |
| injured              | Number of injured, non-fatal victims in the incident, excluding the shooter |
| total_victims        | number of total victims in the incident, excluding the shooter              |
| location_type        | generic location in which the shooting took place                           |
| male                 | logical value, indicating whether the shooter was male                      |
| age_of_shooter       | age of the shooter when the incident occured                                |
| race                 | race of the shooter                                                         |
| prior_mental_illness | did the shooter show evidence of mental illness prior to the incident?      |


-   Generate a data frame that summarizes the number of mass shootings per year.

```{r}
yearly_shootings_summary <- mass_shootings %>% 
  group_by(year) %>% 
  summarise(shootings=n()) %>% 
  arrange(year)

print(yearly_shootings_summary)
```

```{r}
library(ggplot2)
# Extract race column. Since we will be converting this to a factor, it is best to create a new dataframe with just the required data
shooter_race <- mass_shootings %>%
  select(race) %>% 
  filter(!is.na(race))

print(shooter_race)

# Convert race column to factor (this this is a categorical variable)
shooter_race$race <- as.factor(shooter_race$race)
# Count the number of mass shooters per race category
shooter_counts <- shooter_race %>%
  group_by(race) %>%
  summarize(Shooters = n()) %>%
  arrange(desc(Shooters))
  
ggplot(shooter_counts, aes(x = reorder(race, -Shooters), y = Shooters)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = Shooters), vjust = -0.5, size = 3, color = "black") +  # Add data labels
  labs(x = "Race", y = "Number of Mass Shooters") +
  ggtitle("Most mass shooters in the US are white") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

-   Generate a boxplot visualizing the number of total victims, by type of location.

```{r}

ggplot(mass_shootings, aes(x = location_type, y = total_victims)) +
  geom_boxplot(fill = "steelblue", color = "black") +
  labs(x = "Location", y = "Total Victims") +
  ggtitle("Number of Total Victims by Location") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# We can see that the outliers distort this graph and undermine its value

```

-   Redraw the same plot, but remove the Las Vegas Strip massacre from the dataset.

```{r}

mass_shootings %>% 
  filter(case != "Las Vegas Strip massacre") %>% 
  ggplot(aes(x = location_type, y = total_victims)) +
  geom_boxplot(fill = "steelblue", color = "black") +
  labs(x = "Location type", y = "Total Victims") +
  ggtitle("Shootings at schools and military bases tend to be the most deadly") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


-   How many white males with prior signs of mental illness initiated a mass shooting after 2000?

```{r}
library(dplyr)

# Filter the dataset for white males with prior signs of mental illness after 2000
white_males_with_mental_issues <- mass_shootings %>%
  filter(male == "TRUE" & race == "White" & prior_mental_illness == "Yes" & year >= 2000, na.rm=TRUE) %>% 
  count(shooter = n())

# Print the result
print(paste("Number of white males with prior signs of mental illness initiating a mass shooting after 2000:", white_males_with_mental_issues))

```

-   Which month of the year has the most mass shootings? Generate a bar chart sorted in chronological (natural) order (Jan-Feb-Mar- etc) to provide evidence of your answer.

```{r}
library(lubridate)

# Count the number of mass shootings per month
shootings_per_month <- mass_shootings %>%
  group_by(month) %>%
  summarise(shootings = n()) %>% 
  mutate(month_num = as.numeric(match(month, month.abb))) %>% 
  mutate(month = fct_reorder(month, month_num, max)) %>%
  arrange(month_num) %>% 
  ggplot(aes(x = month, y = shootings)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Month", y = "Number of Mass Shootings") +
  ggtitle("Most mass shootings occur in February") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(shootings_per_month)

```

-   How does the distribution of mass shooting fatalities differ between White and Black shooters? What about White and Latino shooters?

```{r}
library(ggplot2)

# Filter the data for white and black shooters
white_black_split <- mass_shootings %>%
  filter(race %in% c("White", "Black") & !is.na(race) & race != "Latino")  %>%
  mutate(month_num = as.numeric(match(month, month.abb))) %>% 
  mutate(month = fct_reorder(month, month_num, max)) %>% 
  select(race, fatalities, month, month_num) 


# Create a bar plot to visualize the distribution
ggplot(white_black_split, aes(x = month, fill = race)) +
  geom_bar(position = "stack", stat = "sum", aes(y = fatalities)) +
  scale_fill_manual(values = c("White" = "blue", "Black" = "red")) +
  labs(x = "Month", y = "Number of Fatalities", fill = "Race") +
  ggtitle("Distribution of Fatalities for White and Black Shootings by Month")

# The bar chart below represents the racial breakdown of shooters between white and black
ggplot(white_black_split, aes(x = month, fill = race)) +
  geom_bar(position = "fill", aes(y = (..count..)/sum(..count..))) +
  scale_fill_manual(values = c("White" = "blue", "Black" = "red")) +
  labs(x = "Month", y = "Percentage", fill = "Race") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggtitle("White shooters generally outnumber black shooters, except in February")


# Recreate the code above, this time filtering for white and latino shooters and removing black shooters.
white_latino_split <- mass_shootings %>%
  filter(race %in% c("White", "Latino") & !is.na(race) & race != "Black")  %>%
  mutate(month_num = as.numeric(match(month, month.abb))) %>% 
  mutate(month = fct_reorder(month, month_num, max))

# Create a bar plot to visualize the distribution
ggplot(white_latino_split, aes(x = month, fill = race)) +
  geom_bar(position = "stack", stat = "sum", aes(y = fatalities)) +
  scale_fill_manual(values = c("White" = "blue", "Latino" = "red")) +
  labs(x = "Month", y = "Number of Fatalities", fill = "Race") +
  ggtitle("Distribution of Fatalities for White and Black Shootings by Month")
```

-   Are mass shootings with shooters suffering from mental illness different from mass shootings with no signs of mental illness in the shooter?

```{r}

# Filter the data for rows with known mental illness status
mass_shootings_with_illness <- mass_shootings[!is.na(mass_shootings$prior_mental_illness), ]

# Reshape the data into long format
reshaped_data <- pivot_longer(mass_shootings_with_illness, cols = c(fatalities, injured, total_victims),
                              names_to = "Variable", values_to = "Count")

print(reshaped_data)

# Create a bar plot facet-wrapped by fatalities, injuries, and total victims
ggplot(reshaped_data, aes(x = prior_mental_illness, y = Count, fill = prior_mental_illness)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge", alpha = 0.5) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(x = "Prior Mental Illness", y = "Mean", fill = "Prior mental illness") +
  ggtitle("On average, shooters with prior-reported mental illnesses are more dangerous")



```

-   Assess the relationship between mental illness and total victims, mental illness and location type, and the intersection of all three variables.

```{r}

# Filter the data for rows with known prior_mental_illness status and location_type
mass_shootings_updated <- mass_shootings[!is.na(mass_shootings$prior_mental_illness) & !is.na(mass_shootings$location_type), ]

# Assess the relationship between prior_mental_illness and total_victims
ggplot(mass_shootings_updated, aes(x = prior_mental_illness, y = total_victims)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Prior Mental Illness", y = "Total Victims") +
  ggtitle("Shooters with mental illnesses tend to claim more victims")

# We can see from this graph that both the average number of victims and variability of victims is higher for shooters with mental illness

# Assess the relationship between prior_mental_illness and location_type
ggplot(mass_shootings_updated, aes(x = prior_mental_illness, fill = location_type)) +
  geom_bar(position = "fill") +
  labs(x = "Prior Mental Illness", y = "Proportion", fill = "Location Type") +
  ggtitle("Mentally ill shooters are more likely to target airports, military bases and religious site")

# We can see here that there is a fairly even distribution of shooting locations among mentally ill and non-mentally ill shooters, however mentally ill shooters are target these other niche locations while non-mentally ill shooters do not 

# Assess the intersection of prior_mental_illness, location_type, and total_victims
ggplot(mass_shootings_updated, aes(x = prior_mental_illness, fill = location_type, y = total_victims)) +
  geom_boxplot() +
  labs(x = "Prior Mental Illness", y = "Total Victims", fill = "Location Type") +
  ggtitle("Mentally ill shooters are more lethal")

# We can see here that mean victims and standard deviations of total victims is higher at all location types for mentally ill shooters as compared with shooters with no prior reported mental illness (with emphasis on "reported"). We can also see that most shooters - either mentally ill or not - tend to cause the most damage at schools.

```

# Exploring credit card fraud


```{r}
#| echo: false
#| message: false
#| warning: false

card_fraud <- read_csv(here::here("data", "card_fraud.csv"))

glimpse(card_fraud)
```

The data dictionary is as follows

| column(variable)      | description                                 |
|-----------------------|---------------------------------------------|
| trans_date_trans_time | Transaction DateTime                        |
| trans_year            | Transaction year                            |
| category              | category of merchant                        |
| amt                   | amount of transaction                       |
| city                  | City of card holder                         |
| state                 | State of card holder                        |
| lat                   | Latitude location of purchase               |
| long                  | Longitude location of purchase              |
| city_pop              | card holder's city population               |
| job                   | job of card holder                          |
| dob                   | date of birth of card holder                |
| merch_lat             | Latitude Location of Merchant               |
| merch_long            | Longitude Location of Merchant              |
| is_fraud              | Whether Transaction is Fraud (1) or Not (0) |

-   In this dataset, how likely are fraudulent transactions? Generate a table that summarizes the number and frequency of fraudulent transactions per year.

```{r}
library(dplyr)
library(tidyr)

view(card_fraud)

# Calculate the likelihood of fraudulent transactions
fraud_likelihood <- card_fraud %>%
  group_by(trans_year,is_fraud) %>% 
  count() %>% 
  group_by(trans_year) %>% 
  mutate(Total_Transactions = sum(n), Percentage_fraudulent = n / Total_Transactions * 100) %>%  
  filter(is_fraud == 1) %>% 
  print

```

-   How much money (in US\$ terms) are fraudulent transactions costing the company? Generate a table that summarizes the total amount of legitimate and fraudulent transactions per year and calculate the % of fraudulent transactions, in US\$ terms.

```{r}

fraud_cost <- card_fraud %>%
  group_by(trans_year) %>% 
  filter(is_fraud == 1) %>%
  summarise(total_loss = paste("$",sum(amt))) %>% 
  print()

# Generate a table summarizing total legitimate and fraudulent transactions per year
transaction_summary <- card_fraud %>%
  group_by(trans_year, is_fraud) %>%
  summarise(total_amount = sum(amt)) %>%
  pivot_wider(names_from = is_fraud, values_from = total_amount, names_prefix = "total_") %>%
  mutate(Percentage_fraud = total_1 / (total_0 + total_1) * 100)

print(transaction_summary)

```

-   Generate a histogram that shows the distribution of amounts charged to credit card, both for legitimate and fraudulent accounts. Also, for both types of transactions, calculate some quick summary statistics.

```{r}

# Filter the dataframe for legitimate and fraudulent transactions separately
legitimate <- card_fraud %>% filter(is_fraud == 0)
fraudulent <- card_fraud %>% filter(is_fraud == 1)

# Create a histogram for legitimate transactions
leg_hist <- ggplot(legitimate, aes(x = amt)) +
  geom_histogram(binwidth = 100) +
  labs(title = "Distribution of Amounts Charged (Legitimate)",
       x = "Amount",
       y = "Frequency")

# Create a histogram for fraudulent transactions
fraud_hist <- ggplot(fraudulent, aes(x = amt)) +
  geom_histogram(binwidth = 100) +
  labs(title = "Distribution of Amounts Charged (Fraudulent)",
       x = "Amount",
       y = "Frequency")

# Calculate summary statistics for legitimate transactions
leg_summary <- legitimate %>% 
  summarise(min_amt = min(amt),
            max_amt = max(amt),
            median_amt = median(amt),
            mean_amt = mean(amt))

# Calculate summary statistics for fraudulent transactions
fraud_summary <- fraudulent %>% 
  summarise(min_amt = min(amt),
            max_amt = max(amt),
            median_amt = median(amt),
            mean_amt = mean(amt))

# Print the summary statistics
print(leg_summary)
print(fraud_summary)

# Display the histograms
print(leg_hist)
print(fraud_hist)



```

-   What types of purchases are most likely to be instances of fraud? Consider category of merchants and produce a bar chart that shows % of total fraudulent transactions sorted in order.

```{r}
# Calculate the percentage of total fraudulent transactions for each category
fraud_category <- fraudulent %>%
  group_by(category) %>%
  summarise(total_fraud = n()) %>%
  mutate(percentage = (total_fraud / sum(total_fraud)) * 100) %>%
  arrange(desc(percentage))

# Create a bar chart showing the percentage of total fraudulent transactions for each category
fraud_chart <- ggplot(fraud_category, aes(x = reorder(category, -percentage), y = percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Percentage of Total Fraudulent Transactions by Merchant Category",
       x = "Merchant Category",
       y = "Percentage") +
  coord_flip()

# Print the data table summarizing the fraudulent transactions by category
print(fraud_category)

# Display the bar chart
print(fraud_chart)
```

-   When is fraud more prevalent? Which days, months, hours? To create new variables to help you in your analysis, we use the `lubridate` package and the following code

```         
mutate(
  date_only = lubridate::date(trans_date_trans_time),
  month_name = lubridate::month(trans_date_trans_time, label=TRUE),
  hour = lubridate::hour(trans_date_trans_time),
  weekday = lubridate::wday(trans_date_trans_time, label = TRUE)
  )
```

-   Are older customers significantly more likely to be victims of credit card fraud? To calculate a customer's age, we use the `lubridate` package and the following code

```         
  mutate(
   age = interval(dob, trans_date_trans_time) / years(1),
    )
```

```{r}
library(lubridate)

time_updated_card_fraud<-card_fraud %>% 
  mutate(
  date_only = lubridate::date(trans_date_trans_time),
  month_name = lubridate::month(trans_date_trans_time, label=TRUE),
  hour = lubridate::hour(trans_date_trans_time),
  weekday = lubridate::wday(trans_date_trans_time, label = TRUE)
  )

view(time_updated_card_fraud)

# Analyse fraud prevalence by day of the week
fraud_by_day <- time_updated_card_fraud %>%
  filter(is_fraud == 1) %>%
  group_by(weekday) %>%
  summarise(total_fraud = n()) %>%
  arrange(desc(total_fraud)) %>% 
  ggplot(aes(x = weekday, y = total_fraud)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Day of the Week", y = "Frequency", title = "Fraud Prevalence by Day of the Week") +
  theme_minimal()

# Analyse fraud prevalence by month
fraud_by_month <- time_updated_card_fraud %>%
  filter(is_fraud == 1) %>%
  group_by(month_name) %>%
  summarise(total_fraud = n()) %>%
  arrange(desc(total_fraud)) %>% 
  ggplot(aes(x = month_name, y = total_fraud)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Month", y = "Frequency", title = "Fraud Prevalence by month") +
  theme_minimal()

# Analyse fraud prevalence by hour
fraud_by_hour <- time_updated_card_fraud %>%
  filter(is_fraud == 1) %>%
  group_by(hour) %>%
  summarise(total_fraud = n()) %>%
  arrange(hour) %>% 
  ggplot(aes(x = hour, y = total_fraud)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Hour of day", y = "Frequency", title = "Fraud Prevalence by Hour") +
  theme_minimal()

# Print the results
print(fraud_by_day)
print(fraud_by_month)
print(fraud_by_hour)

```


# Deliverables

There is a lot of explanatory text, comments, etc. You do not need these, so delete them and produce a stand-alone document that you could share with someone. Knit the edited and completed R Markdown (qmd) file as a Word or HTML document (use the "Knit" button at the top of the script editor window) and upload it to Canvas. You must be comitting and pushing your changes to your own Github repo as you go along.

# Details

-   Who did you collaborate with: Saagar Hemrajani, Chat GPT
-   Approximately how much time did you spend on this problem set: 6 hours
-   What, if anything, gave you the most trouble: The length and scope of the questions.



> As a true test to yourself, do you understand the code you submitted and are you able to explain it to someone else? 

<!-- For the most part, yes. But there were certainly points where I borrowed sections that I didn't fully understand syntax-wise and pasted them into my code to make something work. --> -->
